D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-14 ootavacuous.tex
Update discussion of volatile
Change the discussion of volatile-object accesses to talk about the
difference between execution in a benign and a non-benign environment.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	\co{volatile} is a hint to the implementation to avoid aggressive
	optimization involving the object because the value of the object
	might be changed by means undetectable by an implementation.
\end{quote}
Following the folklore, we interpret these comments to mean that
volatile accesses should be treated as a form of I/O,
implying that volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which must return the value of the store they read from).
As a corollary, the rfe relation does not apply to volatile loads
and hence the accesses in an OOTA cycle must not be volatile.
We also understand that a compiler may not
invent, omit, merge, or reorder volatile accesses,
in a sense to be discussed in
section~\ref{sec:Volatile and Quasi Volatile Accesses} below.

@@

To express this idea in more formal terms, we will augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores).
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile
accesses are considered to be a form of I/O.

For accesses to quasi volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to quasi volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

>>>>

	\co{volatile} is a hint to the implementation to avoid aggressive
	optimization involving the object because the value of the object
	might be changed by means undetectable by an implementation.
\end{quote}

Taking our cue from the folklore, we propose to recognize formally
that programs with volatile objects can execute in two different kinds
of environment: a benign one in which accesses to these objects work
the same as non-volatile memory accesses, and a non-benign one in which
accesses to volatile objects are subject to outside interference and
act more like I/O.
In particular, when running in a non-benign environment a program's
volatile loads can return unpredictable values.
They don't necessarily read from stores (in contrast to non-volatile loads,
which always must return the value of the store they read from).
This implies that volatile load-acquire's do not synchronize with
volatile store-releases in the sense of the C++ memory model,
so they do not contribute in the same way to the happens-before relation.
Also, in these environments the rfe relation does not apply
to volatile loads and hence the accesses in an OOTA cycle must be
non-volatile.

Of course, the machine-code file produced by a compiler must work
properly in either kind of environment.
Therefore the compiler must generally treat accesses to volatile objects
as a form of I/O, and it may not
invent, omit, merge, or reorder these accesses, as we will discuss in
section~\ref{sec:Volatile and Quasi Volatile Accesses} below.

@@

To express this idea in more formal terms,
and to explain what we mean by ``quasi-volatile'' object accesses,
we augment the
requirements for a hardware execution $H$ to realize an abstract
execution $A$.
Each realization must include a map from the set of accesses of
volatile objects in $A$ to the set of instructions in $H$ that access
these objects, having the following properties:
\begin{itemize}
\item	The map connects accesses of the same type (loads to loads
	and stores to stores) and to the same object.
\item	The map connects accesses in a thread of $A$ to accesses in
	the corresponding thread of $H$.
\item	The map is value-preserving: The value of an access in $A$ must be
	the same as the value of the access it maps to in $H$.
\item	In benign environments the map must preserve the rf relation.
	That is, if volatile load $R$ in $A$ reads from store $W$ then
	then the instruction it maps to in $H$ must read from the
	instruction that $W$ maps to.
\item	The map is order-preserving: Two accesses in the same thread
	of $A$ must map to accesses occurring in the same order in $H$.
	(In other words, the compiler may not reorder accesses
	to volatile objects.)
\item	The map is onto: For every access in $H$ to a volatile object
	there must be an access in $A$ that maps to it.
	(In other words, the compiler may not invent accesses to
	volatile objects.)
\item	The map is one-to-one: Different accesses in $A$ map to
	different accesses in $H$.
	(In other words, the compiler may not merge accesses to
	volatile objects.)
\item	The map is total: Every access in $A$ to a volatile object
	maps to some access in $H$.
	(In other words, the compiler may not omit accesses to
	volatile objects.)
\end{itemize}
Most of these are direct consequences of the fact that volatile-object
accesses are considered to be a form of I/O when the program runs
in a non-benign environment.
But to be clear, these requirements apply to all environments.

By contrast, accesses to quasi-volatile objects are normal memory
accesses, not subject to unpredictable interference in non-benign
environments (otherwise the program's behavior would be undefined).
However, we do impose most of the requirements above on quasi-volatile
object accesses.
The last two items are left out: Compilers are allowed to merge or
omit accesses to these objects.
Because of this, the item about preserving the rf relation
applies only when $R$ is not omitted, in which case $W$ must not
be omitted either, but now it applies in all environments.
Lastly, the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same quasi-volatile object.
Accesses to different objects may be reordered relative to each other.
